% !TEX root=maha-nlp-notes.tex
\documentclass{report}
\input{preamble}
\input{macros}
\input{letterfonts}

\title{\textsf{ \huge INTRODUCTION TO} \\\huge{{\bf NATURAL LANGUAGE PROCESSING}}\\\textsf{\large CLASS NOTES}}
\author{\huge{\bf \texttt{ Mahanth Yalla}}\\ \texttt{\large M. Tech-AI, IISc}} 
\date{}

\begin{document}
\maketitle
\newpage
\pdfbookmark[section]{\contentsname}{toc}

\section*{Preface}
These notes are based on the lectures delivered by {\bf Prof. Danish Pruthi} in the course {\bf DS 207 - Introduction to Natural Language Processing} 
at Indian Institute of Science (IISc) Bengaluru - Jan Semester 2025. 
The notes are intended to be a concise summary of the lectures and are not meant to be a replacement for the lectures. 
The notes are written in \LaTeX\ . 

\section*{Disclaimer}
These notes are not official and may contain errors. Please refer to the official course material for accurate information.

\textit{"I cannot guranatee the correctness of these notes. Please use them at your own risk".}

\hfill- Mahanth Yalla 

\subsection*{Contribution}
If you find any errors or have any suggestions, please feel free to open an issue or a pull request on this GitHub repository, I will be happy to incorporate them.
\subsection*{Feedback}
If you have any feedback or suggestions on the notes, please feel free to reach out to me via social media or through mail  \text{mahanthyalla [at] \{iisc [dot] ac [dot] in , gmail [dot] com\}}.

\vfill
{\bf Acknowledgements:} I would like to thank Prof. Danish Pruthi for delivering the lectures and providing the course material. I would also like to thank the TAs for their help and support.


\tableofcontents
\pagebreak


\end{document}
